Estimation of the perceived time of presence of sources in urban acoustic environments using deep learning techniques

### 2019-09-18 paper/paperV2.tex
ml proof v2

The impact of urban sound on human beings has often been studied from a negative point of view (sound pollution), but in the two last decades with the soundscape approach, the positive impact has been revealed (resourcing spaces). The literature shows that the recognition of sources plays a great role in the way humans are affected by sound environments. There is thus a need for characterizing urban acoustic environments not only with sound pressure measurements but also with source-specific attributes such as their perceived time of presence.

This paper demonstrates, on a controlled dataset, that machine learning techniques based on state of the art neural architectures can predict the perceived time of presence of several sound sources. To do so, a corpus of simulated sound scenes is first designed. Perceptual attributes corresponding to those stimuli are gathered through a listening experiment. The availability of the contributions of the individual sound sources within the simulated sound scenes allows the computation of a physical indicator approximating the perceived time of presence of sources, then used for the training and validation of the proposed estimation procedure. The resulting model predicts the time of presence with excellent performance, which translates to the estimation of pleasantness in urban sound environments at a sufficient degree of precision.

@@

Leveraging the advent of the Internet of Things (IoT) and the availability of low-cost sensor networks~\cite{ardouin2018, mydlarz2017} could allow us to characterize the sound environment in a way that is closer to the perception of city dwellers~\cite{iso2014}, by identifying its composition in terms of sources of interest from acoustic measurements at a reasonable cost. Classical monitoring applications mainly rely on the measurement of sound levels on time scales from 1s to several hours~\cite{can2008, brocolini2013}, which offers limited information regarding the content of the sound environment. Conversely, sound environment recognition and event detection applications operate on spectral representations of recorded audio on much finer time scales, in tens of milliseconds~\cite{lunden2016, aucouturier2007, cakir2015}. While using such information-rich representations of the signal also allows the computation of commonly used monitoring acoustical indicators, it may not be desirable in long-term monitoring applications where analysis is performed offline. Indeed, it requires large transmission bandwitdh and storage capabilities, and raises privacy concerns if intelligible speech can be retrieved. Some monitoring applications~\cite{aumond2017, torija2013, nilsson2007} use third octave sound level fast (125ms) measurements to underline relevant spectral information. This representation is easier to store in the long term and reduces privacy concerns~\cite{gontier2017}. In~\cite{aumond2017}, the authors successfully linked acoustical indicators derived from third octave sound levels to the perceived activity of specific sound sources in recordings of polyphonic sound scenes. Though the discriminative properties of the proposed indicators are limited and the underlying methodology cannot be easily extended to larger sound source taxonomiesn it demonstrates that the information content of third ocrave spectra is sufficient to identify sound sources.

The characterization of urban soundscapes through standardized perceptual descriptors has been extensively studied~\cite{viollon2000, axelsson2010, cain2013, jeon2018, aletta2016}. A two-dimensional soundscape model of perceived quality emerges where one dimension corresponds to the pleasantness of the soundscape, and the orthogonal one corresponds to its enventfulness~\cite{axelsson2010, aumond2017, delaitre2014}. Three major source types are found to contribute to these dimensions in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010}, though the exact taxonomy of sources used differs across studies~\cite{guastavino2007, gygi2007, brown2011}. Furthermore, traffic noise, human voices and birds calls can respectively be used as a proxy of these source types~\cite{lavandier2006, ricciardi2014, aumond2017}. In existing models, source activity can be quantified by the dominance~\cite{hong2017, axelsson2010} or the time of presence~\cite{ricciardi2014}. The perceived time of presence presents the advantage of considering all simultaneously active sources, and its identification from acoustic monitoring measurements would provide more complete information on the content of polyphonic urban environments~\cite{lavandier2016}.

In this paper, we do so by developing new indicators relying on source recognition models based on deep learning techniques, which is a growing interest in the machine listening community~\cite{mesaros2017}. In the context of urban sound environments, deep learning has successfully been applied to sound event detection~\cite{foggia2015, mcfee2018, adavanne2017} and sound scene classification~\cite{valenti2016, salamon2017-2}. Several architectures may be used to such aim, the most common of which are convolutional and recurrent neural networks~\cite{mesaros2019}. Learning-based approaches mainly process spectral representations of the audio, such as Mel spectrograms or Mel frequency cepstrum coefficients~\cite{barchiesi2015}. Large amounts of data with task-specific annotations are required for deep learning architectures to learn to extract relevant information from recordings~\cite{salamon2014, mesaros2017, gemmeke2017}. To the best of our knowledge such databases do not currently exist in the literature in the case of source recognition with labels consisting of perceived source activity. It is possible to manually annotate each recording in the corpus. However, this process can be time-consuming, and must be repeated to extend the taxonomy of relevant sound sources. As an alternative, we consider the use of simulated corpora~\cite{lafay2016, salamon2017}, of which the generation procedure provides complete knowledge about the composition of sound scenes. By using a simple indicator computed on this information-rich data to approximate the source-specific perceived time of presence labels, the excerpts composing large corpora may be automatically annotated without the need for additional subjective inputs.

More precisely, the contributions of this paper are three-fold:
\begin{enumerate}
  \item Provide a corpus of simulated scenes for which relevant acoustic properties are well controlled and perceptual judgements are available\footnote{Corpus available at \url{https://zenodo.org/record/3248734\#.XQjC4v7gqUk}}.
  \item Propose numerical means\footnote{Open source code is available at \url{https://github.com/felixgontier/soundSourcePresenceEstimation}.} for predicting the perceived time of presence of sound sources from raw acoustic data using deep learning approaches, with a larger corpus of simulated scenes\footnote{Corpus available at \url{https://zenodo.org/record/3248703\#.XQjDVv7gqUk}}.
  \item Demonstrate through a comparative study with state of the art approaches that the proposed method can be applied to the prediction of a perceptual descriptor of soundscapes to a sufficient degree of accuracy.
\end{enumerate}

First, a procedure for simulating sound scenes corpora is designed in Section~\ref{sec:data} and a listening test is conducted to justify its relevance in the context of a study on perceived quality. A large corpus is constructed and used in Section~\ref{sec:pred} to train and evaluate a deep learning architecture that performs multi-label sound source recognition at relevant time scales. The resulting time of presence are further applied in Section~\ref{sec:app} to the prediction of pleasantness in urban environments, in comparison to existing models from perceptual and physical indicators.

@@

Adjustments are made on the variance of the considered properties to better cover possible scenarios. Furthermore, since voice events in the isolated source database consist mostly of read English recordings, voice event mean EBR are reduced to improve realism. An additional \textit{square} ambiance is added with properties derived empirically from available data. Diverse new scenarios are then generated by sampling the background and event properties distributions and used with random samples from the database of isolated occurences to simulate sound scenes. At the end of the simulation process each sound scene is associated to a random sound level, which is conditioned to the ambiance according to typical sound levels in urban context.

@@

An experiment in the form a listening test is conducted in order to validate the perceptual correspondance of the proposed corpus with urban sound scene recordings, as well as obtain subjective annotations for the design and evaluation of a time of presence estimation model.

Given the aim of this test, a separate corpus is constructed. First, 200 simulated scenes of 45s each with diverse new scenarios are generated using the procedure described in Section~\ref{sec:data_corp}, though with a separate database of isolated source occurences. These scenes are equally distributed among the five considered ambiances (resp. \textit{park}, \textit{quiet street}, \textit{noisy street}, \textit{very noisy street} and \textit{square}). For each scene the perceived time of presence for the traffic, voice and bird sources is estimated using the indicators proposed in~\cite{gontier2018}. Ideally, sound scenes should cover the resulting 3-dimensional space in a homogeneous way. To do so, the 75 scenes that maximize the minimum pairwise distance in the 3-dimensional space are selected. The distribution of selected extracts is shown in Figure~\ref{fig:tvb_pres}. Playback sound levels are realistically drawn as described in Section~\ref{sec:data_corp} and range from 46.6~dB SPL to 77.1~dB SPL over the 75 sound scenes.

@@

The considered indicators are compared to subjective assessments obtained during the listening test to identify their capacity to explain dimensions of soundscape perception. The analysis is performed using arithmetic mean of subjective assessments over all participants for replicated and simulated scenes. The source-specific sound level $L_{eq, s}$, emergence $\Delta L_{s}$ and estimated time of presence $\hat T_s(\alpha, \beta)$ indicators cannot be directly computed on the 6 recorded scenes, as their ground truth source composition is unknown. Thus, these 6 scenes are excluded of the study. Two simulated scenes containing only one source are also excluded because of implementation concerns, resulting in $n=92$ scenes.

@@

\section{Source time of presence prediction}
\label{sec:pred}

\subsection{Deep learning model for presence prediction}
\label{sec:pred_deep}

@@

The performance of the source detection deep learning architecture presented in Section~\ref{sec:pred_deep} on the evaluation dataset is summarized in Table~\ref{tab:perf_cmp}. The overall presence detection accuracy is 92.11\%. The model performs similarly well for the three sources, ranging from 90.08\% for birds to 94.09\% for traffic. Regarding the type of errors, they are equally split between false positives and false negatives. Though, these rates vary depending on the type of source. Traffic has a very low false positive rate at 2.46\% and high false negative rate, while birds display the highest false positive rate at 12.30\%. This is expected given the spectral content related to these sources. Traffic is usually located in lower frequencies than voice or bird events, but may contain high frequency components mistaken for birds by the model. The resulting error on the time of presence is 0.12 RMSE overall on a 0-1 scale, and is close for the three sources.

@@

In the context of urban soundscape quality assessment, pleasantness emerges as the first affective dimension~\cite{axelsson2010}. Pleasantness has been modeled from both the perceived overall loudness in a scene~\cite{blauert1997, jekosch2004} and its content in terms of active sources~\cite{axelsson2010, lavandier2006, ricciardi2014, aumond2017}. In this context, the proposed method for predicting the time of presence of sources is applied to pleasantness estimation.

@@

Baseline models of pleasantness are constructed using data from the experiment in Section~\ref{sec:data_exp}. We first study the relations between perceptual scales with respect to existing work. Table~\ref{tab:percc} shows the Pearson's correlation coefficients between pairs of parameters with assessments averaged for each scene (n=100). The resulting values are consistent with the literature~\cite{aumond2017, gontier2018}, both between general scales as discussed in Section~\ref{sec:data_val} and for the relations to source contributions. Indeed, pleasantness (P) is mainly influenced negatively by overall loudness (OL) and traffic and positively by birds. In previous studies a small positive contribution of voices to pleasantness was found, while no direct relation is visible from the data gathered. This can be explained by the choice of speech samples used during generation, which consist of read audiobooks extracts and thus may sound unnatural in the considered urban environments. Liveliness (L) is here negatively though weakly (r=-0.36) correlated with the presence of birds, while no significant relation was found in previous studies. However, its correlation with the presence of human voices is the strongest (r=0.60), as suggested by~\cite{axelsson2010}. Relations between source-specific parameters are also weak with the exception of traffic being negatively correlated with birds (r=-0.42). The perceived sound level of passing vehicles $L_{T, p}$ is a potential alternative to the time of presence of traffic $T_{T, p}$, as it displays higher correlations with general scales and lower correlations with other source-specific parameters.

@@

The proposed deep learning method is applied to the perceptual experiment corpus to obtain estimations for the time of presence of traffic, voices and birds. These estimations are then applied to pleasantness estimation and compared to models presented in Section~\ref{sec:app_mdls}. To evaluate the model's robustness to the increased polyphony and source complexity of scenes in real-life scenarios, the listening test corpus is split in three parts: the 6 recorded scenes which contain additional sources not present in the datasets simulated for the optimization of the deep learning model, the 19 replicated scenes that also include additional sources as annotated in~\cite{gloaguen2017}, and the 75 simulated scenes that are obtained from the same simulation process as both the development and evaluation dataset.

Pleasantness predictions are obtained by substituting time of presence estimates computed from the presence detection architecture's outputs to the $\hat P_{1, \varphi}$ model presented in Section~\ref{sec:app_mdls}. Thus, only the outputs corresponding to the presence of birds are used. The results are compared to estimates using ground truth $\hat T_s(\alpha_{opt}, \beta_{opt})$ labels in Table~\ref{tab:pppred}. First, pleasantness estimations are equally effective using the model's predictions or the ground truth labels, with about 0.84 RMSE and 82\% of explained variance on the perceptual experiment corpus. This performance of the detection model is expected given the low errors on time of presence estimates in Table~\ref{tab:perf_cmp}. Labels from the detection model result in lower errors on the first sub-corpus containing replicated scenes with sources not seen during training. This result may indicate that the detection model generalizes well to additional sources, though a larger sample size would be required to confirm this interpretation. For the simulated subcorpus of the experiments all performance metrics are comparable. Applying the detection model on the corpus of recorded scenes for which ground truth pleasantness is available results in a RMSE of 1.09 and a decrease in $R^2_{adj}$. This is expected as these scenes are the most distant from the training corpus in terms of sources and scenarios.

@@

Though, the performance of the trained model and its capacity to generalize to recorded data rely on several conditions. First, the perceptual characteristics of the simulated corpus should be comparable to that of typical urban environments. According to the results of the listening test presented in Section~\ref{sec:data_val}, the simulation process of sound scenes does not significantly affect the relations between abstract or content-related subjective descriptors, even with newly generated scenarios and a reduced source taxonomy. This is further demonstrated by the model of pleasantness found in Section~\ref{sec:app_mdls} from subjective annotations of source activity in this experiment. The perceptual models on the listening test corpus are similar to those found in previous studies with both textit{in situ} questionnaires~\cite{aumond2017} and laboratory experiments using recordings~\cite{ricciardi2014}. Second, the quality of predictions is bound by that of annotations. Here, the proposed indicator does not perfectly correspond with the perceived time of presence obtained during the listening test. Furthermore, the indicator is computed using two parameters optimized on available subjective data, though it was found to be quite robust by using a cross-validation scheme during its optimization. While the quality of predictions from the model are encouraging, testing on a larger corpus is required to fully assess its generalization capabilities.

Provided enough excerpts of isolated source occurences are available, the methodology proposed in this paper can be extended to include additional sound sources as the scene simulation process, the time of presence indicator and the deep learning model are all independant from the considered taxonomy. Of course, the complexity of the learning process scales with the number of considered simultaneously active sources. Particularly, differentiating between sources with similar spectral shapes may yield higher error rates as the deep learning architecture relies on the identification of patterns in third octave spectra. To obtain sufficient amounts of data, data augmentation techniques can be used to diversify the training dataset by applying slight modifications to existing examples, such as filtering, pitch shifting or time stretching.

Future work will focus on studying the robustness of the proposed prediction scheme to a refined sound sources taxonomy as well as its application to a large scale sensor network.

>>>>

The impact of urban sound on human beings has often been extensively studied from a negative point of view (sound pollution). In the two last decades, the interest of studying its positive impact has been revealed with the soundscape approach (resourcing spaces). The literature shows that the recognition of sources plays a great role in the way humans are affected by sound environments. There is thus a need for characterizing urban acoustic environments not only with sound pressure measurements but also with source-specific attributes such as their perceived time of presence.

This paper demonstrates, on a controlled dataset, that machine learning techniques based on state of the art neural architectures can predict the perceived time of presence of several sound sources at a sufficient accuracy. To validate this assertion, a corpus of simulated sound scenes is first designed. Perceptual attributes corresponding to those stimuli are gathered through a listening experiment. The availability of the contributions of the individual sound sources within the simulated sound scenes allows the computation of a physical indicator approximating the perceived time of presence of sources, then used for the training and validation of the proposed estimation procedure. The resulting model predicts the time of presence with excellent performance, which translates to the estimation of pleasantness in urban sound environments at a sufficient degree of precision.

@@

Leveraging the advent of the Internet of Things (IoT) and the availability of low-cost sensor networks~\cite{ardouin2018, mydlarz2017} could allow us to characterize the sound environment in a way that is closer to the perception of city dwellers~\cite{iso2014}. This requires the identification of its composition in terms of sources of interest from acoustic measurements at a reasonable cost. Most monitoring applications mainly rely on the measurement of sound levels on time scales from 1s to several hours~\cite{can2008, brocolini2013}, which offers limited information regarding the content of the sound environment. Conversely, sound environment recognition and event detection applications operate on spectral representations of recorded audio on much finer time scales, in tens of milliseconds~\cite{lunden2016, aucouturier2007, cakir2015}. While using such information-rich representations of the signal also allows the computation of commonly used monitoring acoustical indicators, it may not be desirable in long-term monitoring applications where analysis is performed offline. Indeed, it requires large transmission bandwidth and storage capabilities, and raises privacy concerns if intelligible speech can be retrieved. Some monitoring applications~\cite{aumond2017, torija2013, nilsson2007} use third octave sound level fast (125ms) measurements to underline relevant spectral information. This representation is easier to store in the long term and reduces privacy concerns~\cite{gontier2017}. In~\cite{aumond2017}, the authors successfully linked acoustical indicators derived from third octave sound levels to the perceived activity of specific sound sources in recordings of polyphonic sound scenes. Though the discriminative properties of the proposed indicators are limited and the underlying methodology cannot be easily extended to larger sound source taxonomies, it demonstrates that the information content of third octave spectra is sufficient to identify sound sources.

The characterization of urban soundscapes through standardized perceptual descriptors has been extensively studied~\cite{viollon2000, axelsson2010, cain2013, jeon2018, aletta2016}. A two-dimensional soundscape model of perceived quality emerges where one dimension corresponds to the pleasantness of the soundscape, and the orthogonal one corresponds to its enventfulness~\cite{axelsson2010, aumond2017, delaitre2014}. Three major source types are found to contribute to these dimensions in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010}, though the exact taxonomy of sources used differs across studies~\cite{guastavino2007, gygi2007, brown2011}. Furthermore, traffic noise, human voices and birds calls can respectively be used as a proxy of these source types~\cite{lavandier2006, ricciardi2014, aumond2017}. In existing models, source activity can be quantified by the dominance~\cite{hong2017, axelsson2010} or the time of presence~\cite{ricciardi2014}. The perceived time of presence has the advantage of considering all simultaneously active sources, and its identification from acoustic monitoring measurements would provide more complete information on the content of polyphonic urban environments~\cite{lavandier2016}.

In this paper, we do so by developing new indicators relying on source recognition models based on deep learning techniques, which has demonstrated state-of-the-art performance in many tasks studied in the machine listening community~\cite{mesaros2017}. In the context of urban sound environments, deep learning has successfully been applied to sound event detection~\cite{foggia2015, mcfee2018, adavanne2017} and sound scene classification~\cite{valenti2016, salamon2017-2}. Several architectures may be used to this aim, the most common of which are convolutional and recurrent neural networks~\cite{mesaros2019}. Learning-based approaches mainly process spectral representations of the audio, such as Mel spectrograms or Mel frequency cepstrum coefficients~\cite{barchiesi2015}. Large amounts of data with task-specific annotations are required for deep learning architectures to learn to extract relevant information from recordings~\cite{salamon2014, mesaros2017, gemmeke2017}. To the best of our knowledge such databases do not currently exist in the literature in the case of source recognition with labels consisting of perceived source activity. It is possible to manually annotate each recording in the corpus. However, this process would be time-consuming, and must be repeated to extend the taxonomy of relevant sound sources. As an alternative, we consider the use of simulated corpora~\cite{lafay2016, salamon2017}, of which the generation procedure provides complete knowledge about the composition of sound scenes. By using a simple indicator computed on this information-rich data to approximate the source-specific perceived time of presence labels, the excerpts composing large corpora may be automatically annotated without the need for additional subjective inputs. Thanks to this data generation procedure, we are able to validate the proposed approach.

More precisely, the contributions of this paper are three-fold:
\begin{enumerate}
  \item Provide a corpus of simulated scenes for which relevant acoustic properties are well controlled and perceptual judgements are available\footnote{Corpus available at \url{https://zenodo.org/record/3248734\#.XQjC4v7gqUk}}.
  \item Propose numerical means\footnote{Open source code is available at \url{https://github.com/felixgontier/soundSourcePresenceEstimation}.} for predicting the perceived time of presence of sound sources from raw acoustic data using deep learning approaches, with a larger corpus of simulated scenes\footnote{Corpus available at \url{https://zenodo.org/record/3248703\#.XQjDVv7gqUk}}.
  \item Demonstrate that the proposed method can be applied to the prediction of a perceptual descriptor of soundscapes to a sufficient degree of accuracy through a comparative study with state of the art approaches.
\end{enumerate}

The procedure for simulating sound scenes corpora is presented in Section~\ref{sec:data} with the outcomes of the listening test conducted to justify the relevance of the sound scenes corpora in the context of a study on perceived quality. A large corpus is constructed and used to train and evaluate a deep learning architecture that performs multi-label sound source recognition at relevant time scales, as presented in Section~\ref{sec:pred}. The application of the resulting time of presence to the prediction of pleasantness in urban environments is then presented in Section~\ref{sec:app}.%The, in comparison to existing models from perceptual and physical indicators.

@@

Adjustments are made on the variance of the considered properties to better cover possible scenarios. Furthermore, since voice events in the isolated source database consist mostly of read English recordings, voice event mean EBR are reduced to improve realism. An additional \textit{square} ambiance is added with properties derived empirically from available data. Diverse new scenarios are then generated by sampling the background and event properties distributions and used with samples randomly selected from the database of isolated occurrences to simulate sound scenes. At the end of the simulation process the sound level of each sound scene is randomly sampled and conditioned to the ambiance according to typical sound levels in urban context.

@@

An experiment in the form a listening test is conducted in order to validate the perceptual correspondence of the proposed corpus with urban sound scene recordings, as well as obtain subjective annotations for the design and evaluation of a time of presence estimation model.

Given the aim of this test, a separate corpus is constructed. First, 200 simulated scenes of 45s each with diverse new scenarios are generated with the procedure described in Section~\ref{sec:data_corp}, though with a separate database of isolated source occurences. These scenes are equally distributed among the five considered ambiances (resp. \textit{park}, \textit{quiet street}, \textit{noisy street}, \textit{very noisy street} and \textit{square}). For each scene the perceived time of presence for the traffic, voice and bird sources is estimated using the indicators proposed in~\cite{gontier2018}. Ideally, sound scenes should cover the resulting 3-dimensional space in a homogeneous way. To do so, the 75 scenes that maximize the minimum pairwise distance in the 3-dimensional space are selected. The distribution of selected extracts is shown in Figure~\ref{fig:tvb_pres}. Playback sound levels are realistically drawn as described in Section~\ref{sec:data_corp} and range from 46.6~dB SPL to 77.1~dB SPL over the 75 sound scenes.

@@

The considered indicators are compared to subjective assessments obtained during the listening test to identify their capacity to explain dimensions of soundscape perception. The analysis is performed using the arithmetic mean of subjective assessments over all participants for replicated and simulated scenes. The source-specific sound level $L_{eq, s}$, emergence $\Delta L_{s}$ and estimated time of presence $\hat T_s(\alpha, \beta)$ indicators cannot be directly computed on the 6 recorded scenes, as their ground truth source composition is unknown. Thus, these 6 scenes are excluded of the study. Two simulated scenes containing only one source are also excluded because of implementation concerns, resulting in $n=92$ scenes.  TODO implementation : more details needed

@@

\section{Prediction of time of presence of sources}
\label{sec:pred}

\subsection{Deep learning for presence prediction}
\label{sec:pred_deep}

@@

The performance of the source detection deep learning architecture presented in Section~\ref{sec:pred_deep} on the evaluation dataset is summarized in Table~\ref{tab:perf_cmp}. The overall presence detection accuracy is 92.11\%. The model performs similarly well for the three sources, ranging from 90.08\% for birds to 94.09\% for traffic. Regarding the type of errors, they are equally split between false positives and false negatives. Though, these rates vary depending on the type of source. Traffic has a very low false positive rate at 2.46\% and high false negative rate, while birds display the highest false positive rate at 12.30\%. This is expected given the spectral content related to these sources. Traffic is usually located in lower frequency bands than voice or bird events, but may contain high frequency components mistaken for birds by the model. The resulting error on the time of presence is 0.12 RMSE overall on a 0-1 scale, and is close for the three sources.

@@

In the context of urban soundscape quality assessment, pleasantness emerges as the first affective dimension~\cite{axelsson2010}. Pleasantness has been modeled from both the perceived overall loudness in a scene~\cite{blauert1997, jekosch2004} and its content in terms of active sources~\cite{axelsson2010, lavandier2006, ricciardi2014, aumond2017}. To demonstrate the potential of the proposed method for predicting the time of presence of sources is applied to pleasantness estimation.

@@

Baseline models of pleasantness are constructed using data from the experiment in Section~\ref{sec:data_exp}. We first study the relationships between perceptual scales with respect to existing work. Table~\ref{tab:percc} shows the Pearson's correlation coefficients between pairs of parameters with assessments averaged for each scene (n=100). The resulting values are consistent with the literature~\cite{aumond2017, gontier2018}, both for the relations to source contributions and between general scales, as discussed in Section~\ref{sec:data_val}. Indeed, pleasantness (P) is mainly influenced negatively by overall loudness (OL) and traffic and positively by birds. In previous studies a small positive contribution of voices to pleasantness was found, while no direct relation is visible from the data gathered in this study. This can be explained by the choice of speech samples used during generation, which consist of read audiobooks extracts and thus may sound unnatural in the considered urban environments. Liveliness (L) is here negatively though weakly (r=-0.36) correlated with the presence of birds, while no significant relation was found in previous studies. However, its correlation with the presence of human voices is the strongest (r=0.60), as suggested by~\cite{axelsson2010}. Relations between source-specific parameters are also weak with the exception of traffic being negatively correlated with birds (r=-0.42). The perceived sound level of passing vehicles $L_{T, p}$ is a potential alternative to the time of presence of traffic $T_{T, p}$, as it displays higher correlations with general scales and lower correlations with other source-specific parameters.

@@

The proposed deep learning method is applied to the perceptual experiment corpus to obtain estimations for the time of presence of traffic, voices and birds. These estimations are then applied to pleasantness estimation and compared to models presented in Section~\ref{sec:app_mdls}. To evaluate the model's robustness to the increased polyphony and source complexity of scenes in real-life scenarios, the listening test corpus is split in three parts: 1) the 6 recorded scenes which contain additional sources not present in the datasets simulated for the optimization of the deep learning model, 2) the 19 replicated scenes that also include additional sources as annotated in~\cite{gloaguen2017}, and 3) the 75 simulated scenes that are obtained from the same simulation process as both the development and evaluation dataset.

Pleasantness predictions are obtained by substituting time of presence estimates computed from the presence detection architecture's outputs to the $\hat P_{1, \varphi}$ model presented in Section~\ref{sec:app_mdls}. Thus, only the outputs corresponding to the presence of birds are used. TODO The results are compared to estimates using ground truth $\hat T_s(\alpha_{opt}, \beta_{opt})$ labels in Table~\ref{tab:pppred} TODO reformuler. First, pleasantness estimations are equally effective using the model's predictions or the ground truth labels, with about 0.84 RMSE and 82\% of explained variance on the perceptual experiment corpus. This performance of the detection model is expected given the low errors on time of presence estimates in Table~\ref{tab:perf_cmp}. Labels from the detection model result in lower errors on the first sub-corpus containing replicated scenes with sources not seen during training. This result may indicate that the detection model generalizes well to additional sources, though a larger sample size would be required to confirm this interpretation. For the simulated subcorpus of the experiments all performance metrics are comparable. Applying the detection model on the corpus of recorded scenes for which ground truth pleasantness is available results in a RMSE of 1.09 and a decrease in $R^2_{adj}$. This is expected as these scenes are the most distant from the training corpus in terms of sources and scenarios.

@@

Though, the performance of the trained model and its capacity to generalize to recorded data rely on several conditions. First, the perceptual characteristics of the simulated corpus should be comparable to that of typical urban environments. According to the results of the listening test presented in Section~\ref{sec:data_val}, the simulation process of sound scenes does not significantly affect the relations between abstract or content-related subjective descriptors, even with newly generated scenarios and a reduced source taxonomy. This is further demonstrated by the model of pleasantness found in Section~\ref{sec:app_mdls} from subjective annotations of source activity in this experiment. The perceptual models on the listening test corpus are similar to those found in previous studies with both \textit{in situ} questionnaires~\cite{aumond2017} and laboratory experiments using recordings~\cite{ricciardi2014}. Second, the quality of predictions is bound by that of annotations. Here, the proposed indicator does not perfectly correspond with the perceived time of presence obtained during the listening test. Furthermore, the indicator is computed using two parameters optimized on available subjective data, though it was found to be quite robust by using a cross-validation scheme during its optimization. While the quality of predictions from the model are encouraging, testing on a larger corpus is required to fully assess its generalization capabilities.

The methodology proposed in this paper can be extended to include additional sound sources provided that enough excerpts of isolated source occurrences are available. Indeed, the scene simulation process, the time of presence indicator, and the deep learning model are all independent from the considered taxonomy. Of course, the complexity of the learning process scales with the number of considered simultaneously active sources. Particularly, differentiating between sources with similar spectral shapes may yield higher error rates as the deep learning architecture relies on the identification of patterns in third octave spectra. To obtain sufficient amounts of data, data augmentation techniques can be used to diversify the training dataset by applying slight modifications to existing examples, such as filtering, pitch shifting or time stretching. TODO cite may be paper of dcase on data augmentation ?

Future work will thus focus on studying the robustness of the proposed prediction scheme to a refined sound sources taxonomy as well as its application to a large scale sensor network.
